\bf The Noisy Work of Uncertainty Visualisation Research: A Review

### 2025-09-06 paper.qmd
corrected a reference and spelling

---
title: "The Noisy Work of Uncertainty Visualisation Research: A Review"
author: 
  - name: Harriet Mason
    url: https://harrietmason.netlify.app/
    orcid: 0009-0007-4568-8215
    email:  harriet.m.mason@gmail.com
    affiliation:
    - name: Monash University
      department: Department of Econometrics and Business Statistics
      city: Melbourne
      country: Australia
  - name: Dianne Cook
    url: https://dicook.org
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu
    affiliation:
    - name: Monash University
      department: Department of Econometrics and Business Statistics
      city: Melbourne
      country: Australia
  - name: Sarah Goodwin
    url: https://www.linkedin.com/in/smgoodwin/
    orcid: 0000-0001-8894-8282
    email: Sarah.Goodwin@monash.edu
    affiliation:
    - name: Monash University
      department: Department of Human Centred Computing
      city: Melbourne
      country: Australia
  - name: Emi Tanaka
    url: https://emitanaka.org/
    orcid: 0000-0002-1455-259X
    email: Emi.Tanaka@anu.edu.au
    affiliation: 
    - name: The Australian National University
      department: Biological Data Science Institute
      city: Canberra
      country: Australia
  - name: Susan VanderPlas
    url: https://srvanderplas.github.io
    orcid: 0000-0002-3803-0972
    email: susan.vanderplas@unl.edu
    affiliation:
    - name: University of Nebraska–Lincoln
      department: Statistics Department
      city:  Lincoln
      country: United States
bibliography: paper.bib
abstract: Uncertainty visualisation is quickly becomming a hot topic in information visualisation. Exisiting reviews in the field take the definition and purpose of an uncertainty visualisation to be self evident which results in a large amout of conflicting information. This conflict largely stems from a conflation between uncertainty visualisations designed for decision making and those designed to prevent false conclusions. We coin the term "signal suppression" to describe a visualisation that is designed for preventing false conclusions, as the approach demands that the signal (i.e. the collective take away of the estimates) is suppressed by the noise (i.e. the variance on those estimates). We argue that the current standards in visualisation suggest that uncertainty visualisations designed for decision making should not be considered uncertainty visualisations at all. Therefore, future work should focus on signal suppression. Effective signal suppression requires us to communicate the signal and the noise as a single "validity of signal" variable, and doing so proves to be difficult with current methods. We illustrate current approaches to uncertainty visualisation by showing how they would change the visual apprearance of a choropleth map. These maps allow us to see why some methods succeed at signal suppression, while others fall short. Evaluating visualisations on how well they perform signal suppression also proves to be difficult, as it involves measuring the effect of noise, a variable we typically try to ignore. We suggest authors use qualitative studies or compare uncertainty visualisations to the relevant hypothesis tests. 
date: last-modified
toc: false
number-sections: true
latex-clean: true
format:
  jasa-pdf:
    keep-tex: true  
    journal:
      blinded: false
  jasa-html: default
fig-valign: bottom
cap-location: bottom
editor_options: 
  chunk_output_type: console
---

@@

# load Libraries
library(tidyverse)
# devtools::install_github("lydialucchesi/Vizumap")
library(Vizumap)
library(RColorBrewer)
library(scales)
library(sf)
library(ggrepel)
# devtools::install_github("UrbanInstitute/urbnmapr")
library(urbnmapr)
library(flextable)
library(colorspace)
library(rgeos)
```

@@

Among these summary tools, information visualisations are one of the most powerful as they allow for quick and memorable communication that allow us to identify quirks in our data that we didn't know to look for.
Datasets such as Anscombe's quartet [@anscombe] or the Datasaurus Dozen [@datasaurpkg] show a case where visual statistics highlight elements of the data that are invisible to the typical summary statistics. 
Something as simple as sketching a distribution before recalling statistics or making predictions can greatly increase the accuracy of those measures [@Hullman2018; @Goldstein2014].

"Uncertainty visualisation" is a relatively new field in research. 
Early mentions of uncertainty visualisation start to appear in the late 1980s [@Ibrekk1987], with geospatial information visualisation literature from the early 1990s declaring this to be essential aspect of any information display [@MacEachren1992; @Carr1992]. 
@fig-ibrekk depicts an example of the uncertainty visualisations discussed in these early papers. 
Despite kicking off the field, these papers did not define uncertainty visualisation.
This has led to a lack of consensus on what it means for a graphic to visualise uncertainty, an issue we will return to later.
Therefore, while the field is considered to be quite new, many of the graphics used for uncertainty visualisation have been around for much longer. 
For example, box plots and histograms display variation which becomes synonymous with uncertainty when we are using them to depict the variation of an estimate. 
Today, there are an abundance of publications on the topic which makes it timely to construct a review of the field.
That is, now that there is an overwhelming amount of information, it is valuable to distil it into simple facts. 
In fact, there have already been several reviews published but a central piece of discussion is missing.

@@

# recurvisely decrease value
v_val = 0.5
bivariatepal <- c(basecols,
                  colsupress(basecols, sat=v_val),
                  colsupress(colsupress(basecols, sat=v_val), sat=v_val),
                  colsupress(colsupress(colsupress(basecols, sat=v_val), sat=v_val), sat=v_val))
# establish levels of palette
names(bivariatepal) <- paste(rep(1:8, 4), "-" , rep(1:4, each=8), sep="")

>>>>

---
title: "The Noisy Work of Uncertainty Visualisation Research: A Review"
author: 
  - name: Harriet Mason
    url: https://harrietmason.netlify.app/
    orcid: 0009-0007-4568-8215
    email:  harriet.m.mason@gmail.com
    affiliation:
    - name: Monash University
      department: Department of Econometrics and Business Statistics
      city: Melbourne
      country: Australia
  - name: Dianne Cook
    url: https://dicook.org
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu
    affiliation:
    - name: Monash University
      department: Department of Econometrics and Business Statistics
      city: Melbourne
      country: Australia
  - name: Sarah Goodwin
    url: https://www.linkedin.com/in/smgoodwin/
    orcid: 0000-0001-8894-8282
    email: Sarah.Goodwin@monash.edu
    affiliation:
    - name: Monash University
      department: Department of Human Centred Computing
      city: Melbourne
      country: Australia
  - name: Emi Tanaka
    url: https://emitanaka.org/
    orcid: 0000-0002-1455-259X
    email: Emi.Tanaka@anu.edu.au
    affiliation: 
    - name: The Australian National University
      department: Biological Data Science Institute
      city: Canberra
      country: Australia
  - name: Susan VanderPlas
    url: https://srvanderplas.github.io
    orcid: 0000-0002-3803-0972
    email: susan.vanderplas@unl.edu
    affiliation:
    - name: University of Nebraska–Lincoln
      department: Statistics Department
      city:  Lincoln
      country: United States
bibliography: paper.bib
abstract: Uncertainty visualisation is quickly becoming a hot topic in information visualisation. Existing reviews in the field take the definition and purpose of an uncertainty visualisation to be self evident which results in a large amout of conflicting information. This conflict largely stems from a conflation between uncertainty visualisations designed for decision making and those designed to prevent false conclusions. We coin the term "signal suppression" to describe a visualisation that is designed for preventing false conclusions, as the approach demands that the signal (i.e. the collective take away of the estimates) is suppressed by the noise (i.e. the variance on those estimates). We argue that the current standards in visualisation suggest that uncertainty visualisations designed for decision making should not be considered uncertainty visualisations at all. Therefore, future work should focus on signal suppression. Effective signal suppression requires us to communicate the signal and the noise as a single "validity of signal" variable, and doing so proves to be difficult with current methods. We illustrate current approaches to uncertainty visualisation by showing how they would change the visual apprearance of a choropleth map. These maps allow us to see why some methods succeed at signal suppression, while others fall short. Evaluating visualisations on how well they perform signal suppression also proves to be difficult, as it involves measuring the effect of noise, a variable we typically try to ignore. We suggest authors use qualitative studies or compare uncertainty visualisations to the relevant hypothesis tests. 
date: last-modified
toc: false
number-sections: true
latex-clean: true
format:
  jasa-pdf:
    keep-tex: true  
    journal:
      blinded: false
  jasa-html: default
fig-valign: bottom
cap-location: bottom
editor_options: 
  chunk_output_type: console
---

@@

# load Libraries
library(tidyverse)
# devtools::install_github("lydialucchesi/Vizumap")
library(Vizumap)
library(RColorBrewer)
library(scales)
library(sf)
library(ggrepel)
# devtools::install_github("UrbanInstitute/urbnmapr")
library(urbnmapr)
library(flextable)
library(colorspace)
#library(rgeos)
```

@@

Among these summary tools, information visualisations are one of the most powerful as they allow for quick and memorable communication that allow us to identify quirks in our data that we didn't know to look for.
Datasets such as Anscombe's quartet [@anscombe] or the Datasaurus Dozen [@datasaurus; @datasaurpkg] show a case where visual statistics highlight elements of the data that are invisible to the typical summary statistics. 
Something as simple as sketching a distribution before recalling statistics or making predictions can greatly increase the accuracy of those measures [@Hullman2018; @Goldstein2014].

"Uncertainty visualisation" is a relatively new field in research. 
Early mentions of uncertainty visualisation start to appear in the late 1980s [@Ibrekk1987], with geospatial information visualisation literature from the early 1990s declaring this to be essential aspect of any information display [@MacEachren1992; @Carr1992]. 
@fig-ibrekk depicts an example of the uncertainty visualisations discussed in these early papers. 
Despite kicking off the field, these papers did not define uncertainty visualisation.
This has led to a lack of consensus on what it means for a graphic to visualise uncertainty, an issue we will return to later.
Therefore, while the field is considered to be quite new, many of the graphics used for uncertainty visualisation have been around for much longer. 
For example, box plots and histograms display variation which becomes synonymous with uncertainty when we are using them to depict the variation of an estimate. 
Today, there are an abundance of publications on the topic which makes it timely to construct a review of the field.
That is, now that there is an overwhelming amount of information, it is valuable to distill it into simple facts. 
In fact, there have already been several reviews published but a central piece of discussion is missing.

@@

# recursively decrease value
v_val = 0.5
bivariatepal <- c(basecols,
                  colsupress(basecols, sat=v_val),
                  colsupress(colsupress(basecols, sat=v_val), sat=v_val),
                  colsupress(colsupress(colsupress(basecols, sat=v_val), sat=v_val), sat=v_val))
# establish levels of palette
names(bivariatepal) <- paste(rep(1:8, 4), "-" , rep(1:4, each=8), sep="")

### 2025-09-06 paper.tex
updated paper with correct spelling and sent to arxiv
corrected a reference and spelling

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt]{article}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{agsm}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The Noisy Work of Uncertainty Visualisation Research: A Review},
  pdfauthor={Harriet Mason; Dianne Cook; Sarah Goodwin; Emi Tanaka; Susan VanderPlas},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\date{November 20, 2024}
\title{\bf The Noisy Work of Uncertainty Visualisation Research: A
Review}
\author{
Harriet Mason\\
Department of Econometrics and Business Statistics, Monash University\\
and\\Dianne Cook\\
Department of Econometrics and Business Statistics, Monash University\\
and\\Sarah Goodwin\\
Department of Human Centred Computing, Monash University\\
and\\Emi Tanaka\\
Biological Data Science Institute, The Australian National University\\
and\\Susan VanderPlas\\
Statistics Department, University of Nebraska--Lincoln\\
}
\maketitle

\bigskip
\bigskip
\begin{abstract}
Uncertainty visualisation is quickly becomming a hot topic in
information visualisation. Exisiting reviews in the field take the
definition and purpose of an uncertainty visualisation to be self
evident which results in a large amout of conflicting information. This
conflict largely stems from a conflation between uncertainty
visualisations designed for decision making and those designed to
prevent false conclusions. We coin the term ``signal suppression'' to
describe a visualisation that is designed for preventing false
conclusions, as the approach demands that the signal (i.e.~the
collective take away of the estimates) is suppressed by the noise
(i.e.~the variance on those estimates). We argue that the current
standards in visualisation suggest that uncertainty visualisations
designed for decision making should not be considered uncertainty
visualisations at all. Therefore, future work should focus on signal
suppression. Effective signal suppression requires us to communicate the
signal and the noise as a single ``validity of signal'' variable, and
doing so proves to be difficult with current methods. We illustrate
current approaches to uncertainty visualisation by showing how they
would change the visual apprearance of a choropleth map. These maps
allow us to see why some methods succeed at signal suppression, while
others fall short. Evaluating visualisations on how well they perform
signal suppression also proves to be difficult, as it involves measuring
the effect of noise, a variable we typically try to ignore. We suggest
authors use qualitative studies or compare uncertainty visualisations to
the relevant hypothesis tests.
\end{abstract}

@@

Among these summary tools, information visualisations are one of the
most powerful as they allow for quick and memorable communication that
allow us to identify quirks in our data that we didn't know to look for.
Datasets such as Anscombe's quartet \citep{anscombe} or the Datasaurus
Dozen \citep{datasaurpkg} show a case where visual statistics highlight
elements of the data that are invisible to the typical summary
statistics. Something as simple as sketching a distribution before
recalling statistics or making predictions can greatly increase the
accuracy of those measures \citep{Hullman2018, Goldstein2014}.

``Uncertainty visualisation'' is a relatively new field in research.
Early mentions of uncertainty visualisation start to appear in the late
1980s \citep{Ibrekk1987}, with geospatial information visualisation
literature from the early 1990s declaring this to be essential aspect of
any information display \citep{MacEachren1992, Carr1992}.
Figure~\ref{fig-ibrekk} depicts an example of the uncertainty
visualisations discussed in these early papers. Despite kicking off the
field, these papers did not define uncertainty visualisation. This has
led to a lack of consensus on what it means for a graphic to visualise
uncertainty, an issue we will return to later. Therefore, while the
field is considered to be quite new, many of the graphics used for
uncertainty visualisation have been around for much longer. For example,
box plots and histograms display variation which becomes synonymous with
uncertainty when we are using them to depict the variation of an
estimate. Today, there are an abundance of publications on the topic
which makes it timely to construct a review of the field. That is, now
that there is an overwhelming amount of information, it is valuable to
distil it into simple facts. In fact, there have already been several
reviews published but a central piece of discussion is missing.

@@

\includegraphics{paper_files/figure-pdf/fig-ibrekk-1.pdf}

@@

\includegraphics{paper_files/figure-pdf/fig-ibrekk-2.pdf}

@@

\includegraphics{paper_files/figure-pdf/fig-ibrekk-3.pdf}

@@

\includegraphics{paper_files/figure-pdf/fig-ibrekk-4.pdf}

@@

\includegraphics{paper_files/figure-pdf/fig-ibrekk-5.pdf}

@@

\includegraphics{paper_files/figure-pdf/fig-ibrekk-6.pdf}

@@

\includegraphics{paper_files/figure-pdf/fig-ibrekk-7.pdf}

@@

\includegraphics{paper_files/figure-pdf/fig-ibrekk-8.pdf}

@@

\includegraphics{paper_files/figure-pdf/fig-ibrekk-9.pdf}

@@

\includegraphics{paper_files/figure-pdf/fig-data-1.pdf}

@@

\includegraphics{paper_files/figure-pdf/fig-data-2.pdf}

@@

\includegraphics{paper_files/figure-pdf/fig-choropleth-1.pdf}

@@

\includegraphics{paper_files/figure-pdf/fig-choropleth-2.pdf}

@@

\includegraphics{paper_files/figure-pdf/fig-choropleth-3.pdf}

@@

\includegraphics{paper_files/figure-pdf/fig-exceed-1.pdf}

@@

\includegraphics{paper_files/figure-pdf/fig-exceed-2.pdf}

@@

\includegraphics{paper_files/figure-pdf/fig-exceed-3.pdf}

@@

\includegraphics{paper_files/figure-pdf/fig-bivariate-1.pdf}

@@

\includegraphics{paper_files/figure-pdf/fig-bivariate-2.pdf}

@@

\includegraphics{paper_files/figure-pdf/fig-bivariate-3.pdf}

@@

\includegraphics{paper_files/figure-pdf/fig-vsup-1.pdf}

@@

\includegraphics{paper_files/figure-pdf/fig-vsup-2.pdf}

@@

\includegraphics{paper_files/figure-pdf/fig-vsup-3.pdf}

@@

\includegraphics{paper_files/figure-pdf/fig-pixel-1.pdf}

@@

\includegraphics{paper_files/figure-pdf/fig-pixel-2.pdf}

@@

\includegraphics{paper_files/figure-pdf/fig-pixel-3.pdf}

>>>>

% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt]{article}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{5}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

\setlength{\emergencystretch}{3em} % prevent overfull lines

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage[]{natbib}
\bibliographystyle{agsm}

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={The Noisy Work of Uncertainty Visualisation Research: A Review},
  pdfauthor={Harriet Mason; Dianne Cook; Sarah Goodwin; Emi Tanaka; Susan VanderPlas},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\date{September 7, 2025}
\title{\bf The Noisy Work of Uncertainty Visualisation Research: A
Review}
\author{
Harriet Mason\\
Department of Econometrics and Business Statistics, Monash University\\
and\\Dianne Cook\\
Department of Econometrics and Business Statistics, Monash University\\
and\\Sarah Goodwin\\
Department of Human Centred Computing, Monash University\\
and\\Emi Tanaka\\
Biological Data Science Institute, The Australian National University\\
and\\Susan VanderPlas\\
Statistics Department, University of Nebraska--Lincoln\\
}
\maketitle

\bigskip
\bigskip
\begin{abstract}
Uncertainty visualisation is quickly becoming a hot topic in information
visualisation. Existing reviews in the field take the definition and
purpose of an uncertainty visualisation to be self evident which results
in a large amout of conflicting information. This conflict largely stems
from a conflation between uncertainty visualisations designed for
decision making and those designed to prevent false conclusions. We coin
the term ``signal suppression'' to describe a visualisation that is
designed for preventing false conclusions, as the approach demands that
the signal (i.e.~the collective take away of the estimates) is
suppressed by the noise (i.e.~the variance on those estimates). We argue
that the current standards in visualisation suggest that uncertainty
visualisations designed for decision making should not be considered
uncertainty visualisations at all. Therefore, future work should focus
on signal suppression. Effective signal suppression requires us to
communicate the signal and the noise as a single ``validity of signal''
variable, and doing so proves to be difficult with current methods. We
illustrate current approaches to uncertainty visualisation by showing
how they would change the visual apprearance of a choropleth map. These
maps allow us to see why some methods succeed at signal suppression,
while others fall short. Evaluating visualisations on how well they
perform signal suppression also proves to be difficult, as it involves
measuring the effect of noise, a variable we typically try to ignore. We
suggest authors use qualitative studies or compare uncertainty
visualisations to the relevant hypothesis tests.
\end{abstract}

@@

Among these summary tools, information visualisations are one of the
most powerful as they allow for quick and memorable communication that
allow us to identify quirks in our data that we didn't know to look for.
Datasets such as Anscombe's quartet \citep{anscombe} or the Datasaurus
Dozen \citep{datasaurus, datasaurpkg} show a case where visual
statistics highlight elements of the data that are invisible to the
typical summary statistics. Something as simple as sketching a
distribution before recalling statistics or making predictions can
greatly increase the accuracy of those measures
\citep{Hullman2018, Goldstein2014}.

``Uncertainty visualisation'' is a relatively new field in research.
Early mentions of uncertainty visualisation start to appear in the late
1980s \citep{Ibrekk1987}, with geospatial information visualisation
literature from the early 1990s declaring this to be essential aspect of
any information display \citep{MacEachren1992, Carr1992}.
Figure~\ref{fig-ibrekk} depicts an example of the uncertainty
visualisations discussed in these early papers. Despite kicking off the
field, these papers did not define uncertainty visualisation. This has
led to a lack of consensus on what it means for a graphic to visualise
uncertainty, an issue we will return to later. Therefore, while the
field is considered to be quite new, many of the graphics used for
uncertainty visualisation have been around for much longer. For example,
box plots and histograms display variation which becomes synonymous with
uncertainty when we are using them to depict the variation of an
estimate. Today, there are an abundance of publications on the topic
which makes it timely to construct a review of the field. That is, now
that there is an overwhelming amount of information, it is valuable to
distill it into simple facts. In fact, there have already been several
reviews published but a central piece of discussion is missing.

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-ibrekk-1.pdf}}

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-ibrekk-2.pdf}}

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-ibrekk-3.pdf}}

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-ibrekk-4.pdf}}

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-ibrekk-5.pdf}}

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-ibrekk-6.pdf}}

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-ibrekk-7.pdf}}

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-ibrekk-8.pdf}}

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-ibrekk-9.pdf}}

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-data-1.pdf}}

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-data-2.pdf}}

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-choropleth-1.pdf}}

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-choropleth-2.pdf}}

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-choropleth-3.pdf}}

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-exceed-1.pdf}}

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-exceed-2.pdf}}

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-exceed-3.pdf}}

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-bivariate-1.pdf}}

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-bivariate-2.pdf}}

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-bivariate-3.pdf}}

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-vsup-1.pdf}}

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-vsup-2.pdf}}

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-vsup-3.pdf}}

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-pixel-1.pdf}}

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-pixel-2.pdf}}

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-pixel-3.pdf}}
